---
title: "Mouse Visium HD, spatial expression and cluster phenotyping"
author: "heinin"
date: "2026-09-09"
output:
  workflowr::wflow_html:
    code_folding: hide
  html_notebook: default
editor_options:
  chunk_output_type: console
---

## Load packages

```{r, warning=F, message=F}

library(Seurat)
library(ggplot2)
library(data.table)
library(dplyr)
library(patchwork)
library(tidyr)
library(ggrepel)
library(tidyverse)
library(googlesheets4)
library(gprofiler2)
library(scCustomize)
library(RColorBrewer)

```

## Environment variables and helper functions

```{r, message = F, warning = F}

set.seed(1234)
options(future.globals.maxSize = 30000 * 1024^2)
options(scipen = 99999)
options(ggrepel.max.overlaps = Inf)

figdir <- "/home/hnatri/Mm_CARSPP1_VisiumHD/figures"

# Project helper functions and themes (create_barplot(), plot themes, colors)
source("/home/hnatri/Mm_CARSPP1_VisiumHD/code/CART_plot_functions.R")

# Seurat cluster identity used throughout the marker / recluster sections. The
# projected clustering at res 1.2 is the starting label: it is reclustered (see
# the "Reclustering" section) into a `new_cluster` label, and from that point
# on `seurat_cluster_var` is switched to "new_cluster" so annotation, markers
# and the TACCO / predicted-ID comparisons all use the new clusters.
seurat_cluster_var <- "projected_snn_res.1.2"

# Assay used for reclustering. The sketch assay holds the sketched cells the
# projected clustering was learned on; fall back to the default assay if absent.
recluster_assay <- "sketch"

```

## Helper functions

```{r, message = F, warning = F}

# Keep only features present in the object's active (or named) assay, so DotPlot
# / FeaturePlot never error on a gene absent from the panel. Marker vectors
# below deliberately include aliases (e.g. "Cd11b", "NKp46") that are not row
# names; those are dropped here.
present_features <- function(sobj, features, assay = NULL) {
  feats <- unique(features)
  if (is.null(assay)) assay <- DefaultAssay(sobj)
  avail <- rownames(sobj[[assay]])
  keep <- feats[feats %in% avail]
  missing <- setdiff(feats, keep)
  if (length(missing))
    message("Dropping genes not in assay '", assay, "': ",
            paste(missing, collapse = ", "))
  keep
}

# Compact DotPlot of `features` across the levels of `group_var`, in the shared
# theme used throughout (classic theme, rotated axis, no axis titles). `flip`
# puts genes on the y-axis (rows) for tall marker panels.
cluster_dotplot <- function(sobj, features, group_var, flip = TRUE,
                            title = NULL, cols = c("gray89", "tomato3")) {
  feats <- present_features(sobj, features)
  p <- DotPlot(sobj, features = feats, group.by = group_var, cols = cols) +
    theme_classic() +
    RotatedAxis() +
    xlab("") + ylab("")
  if (flip) p <- p + coord_flip()
  if (!is.null(title)) p <- p + ggtitle(title)
  p
}

# Full re-process of one Seurat cluster: subset, then re-run
# FindVariableFeatures / ScaleData / PCA / UMAP / neighbors / clustering from
# scratch on the sketch assay, producing a fresh "umap.sketch" reduction and a
# `subcluster` label (parentCluster_subID). Heavier than reclustering on the
# existing embedding, but gives clean, cluster-specific subclusters.
recluster_subset <- function(sobj, cl, cluster_col = seurat_cluster_var,
                             assay = recluster_assay, npcs = 30, dims = 1:30,
                             res = 0.5) {
  cells <- colnames(sobj)[!is.na(sobj@meta.data[[cluster_col]]) &
                            sobj@meta.data[[cluster_col]] == cl]
  sub <- subset(sobj, cells = cells)

  DefaultAssay(sub) <- if (assay %in% Assays(sub)) assay else DefaultAssay(sub)

  sub <- FindVariableFeatures(sub, verbose = FALSE)
  sub <- ScaleData(sub, verbose = FALSE)
  sub <- RunPCA(sub, npcs = npcs, reduction.name = "pca.sub", verbose = FALSE)
  sub <- FindNeighbors(sub, reduction = "pca.sub", dims = dims, verbose = FALSE)
  sub <- FindClusters(sub, resolution = res, verbose = FALSE)
  sub <- RunUMAP(sub, reduction = "pca.sub", dims = dims,
                 reduction.name = "umap.sketch",
                 reduction.key = "umapsketch_", verbose = FALSE)

  sub$subcluster <- paste0(cl, "_", as.character(sub$seurat_clusters))
  sub$subcluster <- factor(sub$subcluster,
                           levels = paste0(cl, "_",
                                           sort(unique(as.integer(as.character(sub$seurat_clusters))))))
  sub
}

```

## Marker gene sets

```{r, message = F, warning = F}

# Mouse GBM / glioma markers
# Human GBM driver genes mapped to mouse orthologs (Mgmt, Idh1, Idh2, Egfr,
# Trp53, Tert), plus commonly used mouse glioma / lineage-state markers.
gbm_genes_hs <- c("MGMT", "IDH1", "IDH2", "EGFR", "PTEN", "TP53", "TERT")
gbm_genes <- tryCatch(
  gorth(gbm_genes_hs, source_organism = "hsapiens",
        target_organism = "mmusculus")$ortholog_name,
  error = function(e) c("Mgmt", "Idh1", "Idh2", "Egfr", "Trp53", "Tert"))

gbm_markers <- unique(c(gbm_genes,
                        "Pten", "Trp53", "Egfr", "Pdgfra",
                        "Olig1", "Olig2", "Sox2", "Gfap", "Nes", "Vim",
                        "Mki67"))

# Mouse immune markers
#  - Total Leukocytes: Ptprc (CD45)
#  - T Cells (Pan): Cd3e, Cd3g, Cd247 (CD3)
#  - Helper T Cells: Cd4
#  - Cytotoxic T Cells: Cd8a, Cd8b1
#  - Regulatory T Cells (Tregs): Foxp3, Cd25 (Il2ra), Ctla4
#  - B Cells: Cd19, Ms4a1 (CD20), Cd79a
#  - NK Cells: Ncr1 (NKp46), Klrb1c (NK1.1), Cd226
#  - Macrophages (Pan): Adgre1 (F4/80), Cd68, Cd11b (Itgam)
#  - M1 Macrophages: Nos2 (iNOS), Cd80, Cd86
#  - M2 Macrophages: Cd163, Mrc1 (CD206), Arg1
#  - cDCs: Itgax (CD11c), H2-Ab1 (MHCII)
#  - pDCs: Siglech, Bst2 (CD317)
#  - Neutrophils: Ly6g, Ly6c, S100a8
#  - Monocytes: Ly6c, Cd11b (Itgam)
#  - HSPCs: Cd34, Kit (c-Kit)
#  Activation / memory: Sell, Ccr7, Cd44, Il7r; exhaustion Pdcd1, Cd274, Havcr2
mouse_immune <- c("Ptprc", "Cd3e", "Cd3g", "Cd247", "Cd4", "Cd8a", "Cd8b1",
                  "Foxp3", "Cd25", "Il2ra", "Ctla4", "Cd19", "Ms4a1", "Cd79a",
                  "Ncr1", "NKp46", "Klrb1c", "Cd226",
                  "Adgre1", "Cd68", "Cd11b", "Itgam",
                  "Nos2", "Cd80", "Cd86",
                  "Cd163", "Mrc1", "Arg1", "Itgax", "Cd11c",
                  "Siglech", "Bst2", "Ly6g", "Ly6c", "S100a8",
                  "Cd34", "Kit",
                  "Sell", "Ccr7", "Cd44", "Il7r",
                  "Pdcd1", "Cd274", "Havcr2")

lymph_markers <- c("Ptprc", "Cd3e", "Cd3g", "Cd247", "Cd4", "Cd8a", "Cd8b1",
                   "Foxp3", "Cd25", "Il2ra", "Ctla4", "Cd19", "Ms4a1", "Cd79a",
                   "Ncr1", "NKp46", "Klrb1c", "Cd226")

# Microglia vs. bone-marrow-derived TAM markers
mg_markers  <- c("P2ry12", "Tmem119", "Cx3cr1", "Hexb", "Siglech",
                 "Sall1", "Selplg", "Csf1r", "Aif1", "Fcrls", "C1qa", "C1qb")
tam_markers <- c("Spp1", "C1qa", "C1qb", "C1qc", "Trem2", "Mrc1", "Cd163",
                 "Apoe", "Gpnmb", "Cd68", "Adgre1", "Itgam", "Tnfrsf12a", "Tnfsf12")

# SPP1
spp1_axis <- c("Spp1", "Jak1", "Jak2")

# Mouse CAF markers
caf_groups <- list(
  "Pan-CAF / identity"      = c("Pdgfra", "Pdgfrb", "Dcn", "Lum", "Vcan",
                                "Sparc", "Thy1", "Fap", "Pdpn", "Col1a1",
                                "Col1a2"),
  "myCAF / contractile"     = c("Acta2", "Tagln", "Tgfb1", "Tgfbi", "Postn",
                                "Myl9", "Myh11", "Tpm1", "Tpm2"),
  "Collagens / ECM"         = c("Col3a1", "Col5a1", "Col5a2", "Col6a1",
                                "Col6a2", "Col6a3", "Col4a1", "Fn1", "Bgn",
                                "Mgp", "Aspn", "Comp", "Sfrp2", "Sfrp4",
                                "Mmp2", "Timp1"),
  "iCAF / inflammatory"     = c("Il6", "Il1b", "Cxcl1", "Cxcl12", "Cxcl14",
                                "Ccl2", "Pdgfra"),
  "apCAF / antigen pres."   = c("Cd74", "H2-Ab1", "H2-Aa", "H2-Eb1"))

caf_markers <- unique(unlist(caf_groups, use.names = FALSE))

# Lineage marker panel for subcluster phenotyping
lineage_groups <- list(
  "Glioma / tumor"    = c("Olig1", "Olig2", "Sox2", "Egfr", "Pdgfra",
                          "Pten", "Trp53", "Mki67", "Vim"),
  "Astrocyte"         = c("Gfap", "Aqp4", "Slc1a3", "Aldh1l1"),
  "Oligodendrocyte"   = c("Mbp", "Plp1", "Mog", "Sox10", "Mag"),
  "Neuron"            = c("Rbfox3", "Snap25", "Syt1", "Meg3"),
  "Microglia"         = c("P2ry12", "Tmem119", "Cx3cr1", "Hexb", "Sall1"),
  "TAM / macrophage"  = c("C1qa", "C1qb", "Aif1", "Csf1r", "Cd68", "Adgre1",
                          "Mrc1", "Spp1"),
  "T / NK cell"       = c("Cd3e", "Cd8a", "Cd4", "Nkg7", "Klrb1c"),
  "B cell"            = c("Cd19", "Ms4a1", "Cd79a"),
  "Fibroblast / CAF"  = c("Col1a1", "Col1a2", "Dcn", "Pdgfra", "Acta2"),
  "Endothelial"       = c("Pecam1", "Cldn5", "Flt1"),
  "Pericyte / mural"  = c("Rgs5", "Pdgfrb", "Notch3", "Cd248"))

lineage_markers <- unique(unlist(lineage_groups, use.names = FALSE))

```

## Import data

```{r, warning=F, message=F}

# The final object with all TACCO labels.
seurat_data <- readRDS("/tgen_labs/banovich/BCTCSF/Mm_VisiumHD/tumor_obj_w_all_tacco_labels_09_09_2026.rds")

# Fixing TMA3 treatment group
seurat_data$Treatment <- ifelse(seurat_data$TMA == 3 & seurat_data$Sample %in% c(166, 159),
                                "Anti-Spp1", seurat_data$Treatment)
seurat_data$Treatment <- gsub(" ", "", seurat_data$Treatment)
seurat_data$Treatment <- gsub("Anti-Spp1/CAR", "Anti-Spp1+CAR", seurat_data$Treatment)
table(seurat_data$Treatment, seurat_data$TMA)

# Treatment_Day, ordered for all downstream grouping.
seurat_data$Treatment_Day <- paste0(seurat_data$Treatment, "_", seurat_data$Day)
seurat_data$Treatment_Day <- factor(seurat_data$Treatment_Day,
                                    levels = c("Tumor_16", "Tumor_20",
                                               "Anti-Spp1_16", "Anti-Spp1_20",
                                               "CAR_16", "CAR_20",
                                               "Anti-Spp1+CAR_16", "Anti-Spp1+CAR_20"))

```

Colors for treatment and day

```{r, warning=F, message=F}

treatment_cols <- c("Anti-Spp1+CAR" = "aquamarine3",
                    "CAR"           = "cyan4",
                    "Anti-Spp1"     = "mediumpurple",
                    "Tumor"         = "deeppink3")

day_cols <- c("16" = "gray40",
              "20" = "azure3")

```

Panel-present gene counts, used to size the marker dot plots (≈0.22 in / gene).

```{r, warning=F, message=F}

n_immune_genes  <- length(present_features(seurat_data, unique(c(lymph_markers, mouse_immune))))
n_gbm_genes     <- length(present_features(seurat_data, gbm_markers))
n_caf_genes     <- length(present_features(seurat_data, caf_markers))
n_lineage_genes <- length(present_features(seurat_data, lineage_markers))

```

## Spatial overview

Treatment and day across the spatial embedding.

```{r, warning=F, message=F, fig.width = 20, fig.height = 8}

p1 <- DimPlot(seurat_data, reduction = "sp", group.by = "Treatment",
              cols = treatment_cols) +
  theme_classic() + coord_fixed() + NoAxes()

p2 <- DimPlot(seurat_data, reduction = "sp", group.by = "Day",
              cols = day_cols) +
  theme_classic() + coord_fixed() + NoAxes()

p1 / p2

pdf(file.path(figdir, "VisiumHD_dimplot_treatment.pdf"), width = 20, height = 8)
p1
dev.off()

pdf(file.path(figdir, "VisiumHD_dimplot_day.pdf"), width = 20, height = 8)
p2
dev.off()

```

Counts and features per spot

```{r, warning=F, message=F, fig.width = 20, fig.height = 12}

FeaturePlot(seurat_data, reduction = "sp",
            features = c("nCount_Spatial", "nFeature_Spatial"),
            ncol = 1, raster = TRUE) &
  theme_classic() & coord_fixed()

pdf(file.path(figdir, "VisiumHD_counts_features.pdf"), width = 20, height = 12)
FeaturePlot(seurat_data, reduction = "sp",
            features = c("nCount_Spatial", "nFeature_Spatial"),
            ncol = 1, raster = TRUE) &
  theme_classic() & coord_fixed()
dev.off()

```

## GBM / glioma gene expression

Mouse orthologs of the human GBM driver panel across the tissue.

```{r, warning=F, message=F, fig.width = 12, fig.height = 24}

FeaturePlot_scCustom(seurat_object = seurat_data,
                     features = present_features(seurat_data, c(gbm_genes, "C1qa")),
                     reduction = "sp", num_columns = 1, raster = FALSE, order = TRUE)

#ggsave(file.path(figdir, "VisiumHD_GBM_genes.png"), width = 12, height = 24)

```

## Tumor microenvironment marker plots

Immune, SPP1 axis and tissue markers

```{r, warning=F, message=F, fig.width = 10, fig.height = 26}

FeaturePlot_scCustom(seurat_object = seurat_data,
                     features = present_features(seurat_data,
                                c("Tigit", "Cd8a", "Cd3e", "Spp1", "Tnfrsf12a",
                                  "Tnfsf12", "C1qa", "Cd44", "Col1a1", "Pten")),
                     reduction = "sp", num_columns = 1,
                     slot = "data", pt.size = 1, na_cutoff = 1, order = TRUE) &
  coord_fixed() & NoAxes()

ggsave(file.path(figdir, "VisiumHD_TME_genes.png"), width = 10, height = 26)

```

SPP1 axis and interferon signaling

```{r, warning=F, message=F, fig.width = 8, fig.height = 12}

FeaturePlot_scCustom(seurat_object = seurat_data,
                     features = present_features(seurat_data, spp1_axis),
                     reduction = "sp", num_columns = 1,
                     slot = "data", pt.size = 1, na_cutoff = 1, order = TRUE) &
  coord_fixed() & NoAxes()

ifn_features <- present_features(seurat_data,
                                 c(spp1_axis, grep("Ifn", rownames(seurat_data), value = TRUE)))

FeaturePlot_scCustom(seurat_object = seurat_data,
                     features = present_features(seurat_data,
                                c("Spp1", "Jak1", "Jak2", "Ifna", "Ifnb1")),
                     reduction = "sp", num_columns = 1,
                     slot = "data", pt.size = 1, na_cutoff = 1, order = TRUE) &
  coord_fixed() & NoAxes()

```

## Marker expression by treatment and day

```{r, warning=F, message=F, fig.width = 8, fig.height = 10}

VlnPlot(seurat_data, features = present_features(seurat_data, spp1_axis),
        split.by = "Treatment", group.by = "Day", pt.size = 0, ncol = 1,
        cols = treatment_cols) &
  theme_classic()

DotPlot(seurat_data,
        features = present_features(seurat_data, c(spp1_axis, grep("Ifn", rownames(seurat_data), value = TRUE))),
        group.by = "Treatment_Day") &
  theme_classic() & coord_flip() & RotatedAxis() & xlab("") & ylab("")

```

Summary dot plot across treatment × day.

```{r, warning=F, message=F, fig.width = 5, fig.height = 3}

dotp <- DotPlot(seurat_data,
        features = present_features(seurat_data,
                   c("Tigit", "Cd8a", "Cd3e", "Spp1", "C1qa", "Cd44", "Col1a1", "Pten")),
        group.by = "Treatment_Day",
        cols = c("gray89", "tomato3")) +
  coord_flip() + theme_classic() + RotatedAxis() + xlab("") + ylab("")
dotp

pdf(file.path(figdir, "VisiumHD_dotplot.pdf"), width = 5, height = 3)
dotp
dev.off()

```

TAM-focused SPP1 axis by treatment × day.

```{r, warning=F, message=F, fig.width = 6, fig.height = 8}

tams <- subset(seurat_data,
               subset = predicted.id_class_immune %in% c("TAM 1", "TAM 2", "prol. TAM"))

DotPlot(tams, features = present_features(tams, spp1_axis),
        group.by = "Treatment_Day") &
  theme_classic()

VlnPlot(tams, features = present_features(tams, spp1_axis),
        group.by = "Treatment_Day", split.by = "predicted.id_class_immune",
        pt.size = 0, ncol = 1, cols = treatment_cols) &
  theme_classic() & RotatedAxis() & xlab("")

```

## Reclustering each Seurat cluster

```{r recluster-all, warning=F, message=F, results = "hide"}

# Base clustering to recluster is always the projected res 1.2 label, regardless
# of what seurat_cluster_var is later switched to.
recluster_base_var <- "projected_snn_res.1.2"

clusters_to_recluster <- setdiff(sort(unique(as.character(seurat_data@meta.data[[recluster_base_var]]))),
                                 "NA")

# Per-cluster reclustering resolution: default 0.5, higher for cluster 2.
recluster_res <- setNames(rep(0.5, length(clusters_to_recluster)), clusters_to_recluster)
if ("2" %in% names(recluster_res)) recluster_res["2"] <- 1.0

# Re-process each cluster once; results reused by the per-cluster chunks below.
recluster_list <- lapply(clusters_to_recluster, function(cl)
  tryCatch(recluster_subset(seurat_data, cl, cluster_col = recluster_base_var,
                            res = recluster_res[[cl]]),
           error = function(e) { message("Recluster failed for ", cl, ": ", conditionMessage(e)); NULL }))
names(recluster_list) <- clusters_to_recluster
recluster_list <- recluster_list[!vapply(recluster_list, is.null, logical(1))]

```

Assigning a new label

```{r build-new-cluster, warning=F, message=F}

# Collect the per-cell subcluster labels (parentCluster_subID) from every
# reclustered subset back onto the full object.
subcluster_map <- setNames(rep(NA_character_, ncol(seurat_data)), colnames(seurat_data))
for (cl in names(recluster_list)) {
  sub <- recluster_list[[cl]]
  subcluster_map[colnames(sub)] <- as.character(sub$subcluster)
}
seurat_data$subcluster <- subcluster_map

# Subclusters to break out into their own clusters. Any spot NOT in this list
# keeps its projected_snn_res.1.2 label ("combine other").
separate_subclusters <- c("0_2",
                          "1_2", "1_3",
                          "10_2", "10_3",
                          "12_2",
                          "14_2",
                          "4_2",
                          "5_2",
                          "6_3",
                          "9_3")

seurat_data$new_cluster <- as.character(seurat_data@meta.data[[recluster_base_var]])
split_idx <- !is.na(seurat_data$subcluster) & seurat_data$subcluster %in% separate_subclusters
seurat_data$new_cluster[split_idx] <- seurat_data$subcluster[split_idx]

# NA handling + ordered factor.
seurat_data$new_cluster <- ifelse(is.na(seurat_data$new_cluster), "NA", seurat_data$new_cluster)
seurat_data$new_cluster <- factor(seurat_data$new_cluster,
                                  levels = sort(unique(seurat_data$new_cluster)))

# Sanity check: how many spots ended up in each (sub)cluster, and confirm every
# requested subcluster was actually found and split out.
print(table(seurat_data$new_cluster))
missing_split <- setdiff(separate_subclusters, unique(seurat_data$subcluster))
if (length(missing_split))
  message("Requested subclusters not found at the current resolution: ",
          paste(missing_split, collapse = ", "))

# Use the new clusters as the primary Seurat cluster identity from here on
# (annotation, markers, TACCO / predicted-ID comparisons).
seurat_cluster_var <- "new_cluster"

# Palette for the new clusters, reused throughout.
seurat_cluster_col <- colorRampPalette(brewer.pal(12, "Paired"))(length(levels(seurat_data$new_cluster)))
names(seurat_cluster_col) <- levels(seurat_data$new_cluster)

```

Subcluster UMAPs

```{r recluster-umaps, warning=F, message=F, fig.width = 12, fig.height = max(4, 3 * ceiling(length(recluster_list) / 3))}

umap_plots <- lapply(names(recluster_list), function(cl) {
  DimPlot(recluster_list[[cl]], reduction = "umap.sketch",
          group.by = "subcluster", label = TRUE, repel = TRUE) +
    theme_classic() + NoLegend() +
    ggtitle(paste0("Cluster ", cl, " subclusters"))
})

wrap_plots(umap_plots, ncol = 3)

```

Lineage markers in the subclusters of each Seurat cluster

```{r recluster-lineage-dot, warning=F, message=F, fig.width = 5, fig.height = max(4, 0.25 * n_lineage_genes)}

for (cl in names(recluster_list)) {
  print(cluster_dotplot(recluster_list[[cl]], lineage_markers,
                        group_var = "subcluster",
                        title = paste0("Lineage markers in cluster ", cl, " subclusters")))
}

```

Immune, GBM and CAF markers within each cluster's subclusters

```{r recluster-marker-panels, warning=F, message=F, fig.width = 8, fig.height = max(4, 0.22 * n_immune_genes)}

for (cl in names(recluster_list)) {
  sub <- recluster_list[[cl]]
  print(cluster_dotplot(sub, unique(c(lymph_markers, mouse_immune)),
                        group_var = "subcluster",
                        title = paste0("Immune markers, cluster ", cl, " subclusters")))
  print(cluster_dotplot(sub, gbm_markers, group_var = "subcluster",
                        title = paste0("GBM markers, cluster ", cl, " subclusters")))
  print(cluster_dotplot(sub, caf_markers, group_var = "subcluster",
                        title = paste0("CAF markers, cluster ", cl, " subclusters")))
}

```

New cluster label on the tissue

```{r new-cluster-sp, warning=F, message=F, fig.width = 10, fig.height = 8}

DimPlot(seurat_data, reduction = "sp", group.by = seurat_cluster_var,
        cols = seurat_cluster_col) +
  theme_classic() + coord_fixed() + NoAxes()

```

## Cell type annotations

```{r, warning=F, message=F}

seurat_data$TACCO_myeloid_label <- ifelse(is.na(seurat_data$TACCO_myeloid_label),
                                          "NA", seurat_data$TACCO_myeloid_label)
seurat_data$predicted.id_tam    <- ifelse(is.na(seurat_data$predicted.id_tam),
                                          "NA", seurat_data$predicted.id_tam)

# The primary Seurat cluster label is now `new_cluster` (built in the
# reclustering section from projected_snn_res.1.2 with selected subclusters
# split out); its factor levels and palette (seurat_cluster_col) are set there.

TACCO_myeloid_label_col <- colorRampPalette(brewer.pal(10, "Paired"))(length(unique(seurat_data$TACCO_myeloid_label)))
names(TACCO_myeloid_label_col) <- sort(unique(seurat_data$TACCO_myeloid_label))

predicted.id_tam_col <- colorRampPalette(brewer.pal(10, "Paired"))(length(unique(seurat_data$predicted.id_tam)))
names(predicted.id_tam_col) <- sort(unique(seurat_data$predicted.id_tam))

```

TACCO myeloid label on the tissue

```{r, warning=F, message=F, fig.width = 10, fig.height = 8}

DimPlot(seurat_data, group.by = "TACCO_myeloid_label", reduction = "sp") &
  coord_fixed()

```

Cell type composition by treatment × day

```{r, warning=F, message=F, fig.width = 8, fig.height = 5}

create_barplot(seurat_data,
               group_var = "Treatment_Day",
               plot_var = "TACCO_myeloid_label",
               plot_levels = sort(unique(seurat_data$TACCO_myeloid_label)),
               group_levels = levels(seurat_data$Treatment_Day),
               plot_colors = TACCO_myeloid_label_col,
               var_names = c("Frequency (%)", ""),
               legend_title = "Cell type")

```

Per-sample proportions with treatment × day distribution

```{r, warning=F, message=F, fig.width = 13, fig.height = 8}

prop_df <- seurat_data@meta.data %>%
  dplyr::count(Sample, Treatment_Day, TACCO_myeloid_label, name = "n") %>%
  dplyr::group_by(Sample) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup() %>%
  tidyr::complete(tidyr::nesting(Sample, Treatment_Day), TACCO_myeloid_label,
                  fill = list(n = 0, prop = 0))

ggplot(prop_df, aes(x = Treatment_Day, y = prop, fill = Treatment_Day)) +
  geom_violin(scale = "width", trim = FALSE, alpha = 0.7) +
  geom_jitter(width = 0.12, size = 0.9, alpha = 0.8) +
  facet_wrap(~ TACCO_myeloid_label, scales = "free_y") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Proportion of cells") +
  theme_bw(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

```

Seurat cluster composition by treatment × day

```{r, warning=F, message=F, fig.width = 8, fig.height = 5}

create_barplot(seurat_data,
               group_var = "Treatment_Day",
               plot_var = seurat_cluster_var,
               plot_levels = sort(unique(seurat_data@meta.data[[seurat_cluster_var]])),
               group_levels = levels(seurat_data$Treatment_Day),
               plot_colors = seurat_cluster_col,
               var_names = c("Frequency (%)", ""),
               legend_title = "Seurat cluster")

```

## Annotation concordance

```{r, warning=F, message=F}

table(seurat_data@meta.data[[seurat_cluster_var]], seurat_data$TACCO_myeloid_label)
table(seurat_data@meta.data[[seurat_cluster_var]], seurat_data$predicted.id_tam)
table(seurat_data$predicted.id_tam, seurat_data$TACCO_myeloid_label)

```

```{r, warning=F, message=F, fig.width = 18, fig.height = 6}

p1 <- create_barplot(seurat_data, group_var = "TACCO_myeloid_label",
                     plot_var = seurat_cluster_var,
                     plot_levels = sort(unique(seurat_data@meta.data[[seurat_cluster_var]])),
                     group_levels = sort(unique(seurat_data$TACCO_myeloid_label)),
                     plot_colors = seurat_cluster_col,
                     var_names = c("Frequency (%)", ""), legend_title = "Seurat cluster")

p2 <- create_barplot(seurat_data, group_var = seurat_cluster_var,
                     plot_var = "TACCO_myeloid_label",
                     plot_levels = sort(unique(seurat_data$TACCO_myeloid_label)),
                     group_levels = sort(unique(seurat_data@meta.data[[seurat_cluster_var]])),
                     plot_colors = TACCO_myeloid_label_col,
                     var_names = c("Frequency (%)", ""), legend_title = "TACCO")

p1 + p2

p3 <- create_barplot(seurat_data, group_var = "predicted.id_tam",
                     plot_var = seurat_cluster_var,
                     plot_levels = sort(unique(seurat_data@meta.data[[seurat_cluster_var]])),
                     group_levels = sort(unique(seurat_data$predicted.id_tam)),
                     plot_colors = seurat_cluster_col,
                     var_names = c("Frequency (%)", ""), legend_title = "Seurat cluster")

p4 <- create_barplot(seurat_data, group_var = seurat_cluster_var,
                     plot_var = "predicted.id_tam",
                     plot_levels = sort(unique(seurat_data$predicted.id_tam)),
                     group_levels = sort(unique(seurat_data@meta.data[[seurat_cluster_var]])),
                     plot_colors = predicted.id_tam_col,
                     var_names = c("Frequency (%)", ""), legend_title = "Seurat label transfer")

p3 + p4

```

## Top marker expression

Top positive markers for each annotation and each Seurat cluster

TACCO labels

```{r, warning=F, message=F, fig.width = 18, fig.height = 6}

seurat_data <- JoinLayers(seurat_data)

Idents(seurat_data) <- seurat_data$TACCO_myeloid_label
tacco_markers <- FindAllMarkers(seurat_data, only.pos = TRUE, return.thresh = 0.1)
tacco_markers_sig <- tacco_markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC)

cluster_dotplot(seurat_data, unique(tacco_markers_sig$gene),
                group_var = "TACCO_myeloid_label", flip = FALSE)

```

Clusters

```{r, warning=F, message=F, fig.width = 20, fig.height = 6}

Idents(seurat_data) <- seurat_data[[seurat_cluster_var, drop = TRUE]]
cluster_markers <- FindAllMarkers(seurat_data, only.pos = TRUE)
cluster_markers_sig <- cluster_markers %>% group_by(cluster) %>% top_n(n = 5, wt = avg_log2FC)

cluster_dotplot(seurat_data, unique(cluster_markers_sig$gene),
                group_var = seurat_cluster_var, flip = FALSE)

# Saving
cluster_markers_more <- cluster_markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)

write.table(cluster_markers_more, "/home/hnatri/VisiumHD_cluster_markers_top10.tsv",
            quote = F, sep = "\t")

cluster_markers_top <- cluster_markers %>% group_by(cluster) %>% top_n(n = 5, wt = avg_log2FC)

cluster_dotplot(seurat_data, unique(cluster_markers_top$gene),
                group_var = seurat_cluster_var, flip = FALSE)

# Saving all sig markers
cluster_markers_sig <- cluster_markers %>% filter(p_val_adj < 0.01 & abs(avg_log2FC) > 1 & pct.1 > 0.1)

table(cluster_markers_sig$cluster)

write.table(cluster_markers_sig, "/home/hnatri/VisiumHD_cluster_markers_sig.tsv",
            quote = F, sep = "\t")

```

Canonical lineage genes across Seurat clusters.

```{r, warning=F, message=F, fig.width = 8, fig.height = 6}

cluster_dotplot(seurat_data,
                c("Olig1", "Olig2", "Idh1", "Egfr", "Pten", "Trp53",
                  "Ptprc", "Col1a1", "Col1a2", "Vim"),
                group_var = seurat_cluster_var, flip = FALSE)

DimPlot(seurat_data, reduction = "sp", group.by = seurat_cluster_var) +
  theme_classic() + coord_fixed() + NoAxes()

```

Per-spot QC by cluster.

```{r, warning=F, message=F, fig.width = 8, fig.height = 12}

qc_cols <- intersect(c("nCount_Spatial", "nFeature_Spatial", "percent.mt"),
                     colnames(seurat_data@meta.data))

seurat_data@meta.data %>%
  group_by(cluster = .data[[seurat_cluster_var]]) %>%
  dplyr::summarise(across(all_of(qc_cols), median), n = dplyr::n()) %>%
  arrange(cluster)

VlnPlot(seurat_data, features = qc_cols, group.by = seurat_cluster_var, pt.size = 0, ncol = 1) &
  theme(axis.text.x = element_text(size = 7))

```

## Immune markers in each Seurat cluster

```{r immune-by-cluster, warning=F, message=F, fig.width = 8, fig.height = max(4, 0.22 * n_immune_genes)}

cluster_dotplot(seurat_data, unique(c(lymph_markers, mouse_immune)),
                group_var = seurat_cluster_var,
                title = "Immune markers by Seurat cluster")

```

Microglia vs. TAM markers by cluster (helps separate resident microglia vs.TAMs).

```{r immune-mg-tam-by-cluster, warning=F, message=F, fig.width = 8, fig.height = max(4, 0.3 * length(present_features(seurat_data, c(mg_markers, tam_markers))))}

cluster_dotplot(seurat_data, unique(c(mg_markers, tam_markers)),
                group_var = seurat_cluster_var,
                title = "Microglia / TAM markers by Seurat cluster")

```

## Mouse GBM markers in each Seurat cluster

```{r gbm-by-cluster, warning=F, message=F, fig.width = 8, fig.height = max(4, 0.3 * n_gbm_genes)}

cluster_dotplot(seurat_data, gbm_markers, group_var = seurat_cluster_var,
                title = "Mouse GBM / glioma markers by Seurat cluster")

```

## CAF markers in each Seurat cluster

CAF program markers (pan-CAF, myCAF, collagens/ECM, iCAF, apCAF)
across Seurat clusters.

```{r caf-by-cluster, warning=F, message=F, fig.width = 8, fig.height = max(4, 0.25 * n_caf_genes)}

cluster_dotplot(seurat_data, caf_markers, group_var = seurat_cluster_var,
                title = "CAF markers by Seurat cluster")

```

## Myeloid expression

Immune / myeloid marker expression by Seurat cluster and by treatment × day,
and top markers per treatment group within the immune clusters.

```{r myeloid-cluster, warning=F, message=F, fig.width = 8, fig.height = max(4, 0.22 * n_immune_genes)}

cluster_dotplot(seurat_data,
                unique(c(lymph_markers, mouse_immune, spp1_axis,
                         grep("Ifn", rownames(seurat_data), value = TRUE))),
                group_var = seurat_cluster_var,
                title = "Immune / myeloid markers by Seurat cluster")

```

TODO: subset clusters

```{r myeloid-treatment, warning=F, message=F, fig.width = 14, fig.height = 5}

immune_clusters <- intersect(c("4", "4_2", "10_2", "12", "12_2", "5", "5_2"),
                             as.character(unique(seurat_data@meta.data[[seurat_cluster_var]])))
immune_only <- subset(seurat_data,
                      cells = colnames(seurat_data)[as.character(seurat_data@meta.data[[seurat_cluster_var]]) %in% immune_clusters])

DotPlot(seurat_data,
        features = present_features(seurat_data, unique(c(lymph_markers, mouse_immune, spp1_axis))),
        group.by = "Treatment_Day") &
  theme_classic() & RotatedAxis() & xlab("") & ylab("")

Idents(immune_only) <- immune_only$Treatment_Day
treatment_day_immune_markers <- FindAllMarkers(immune_only, only.pos = TRUE)
treatment_day_immune_markers_top <- treatment_day_immune_markers %>%
  group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)

DotPlot(immune_only, features = unique(treatment_day_immune_markers_top$gene),
        group.by = "Treatment_Day") &
  theme_classic() & RotatedAxis() & xlab("") & ylab("")

```

## CAF expression

CAF program markers by Seurat cluster and by treatment × day.

```{r caf-expression, warning=F, message=F, fig.width = 8, fig.height = max(4, 0.25 * n_caf_genes)}

cluster_dotplot(seurat_data, caf_markers, group_var = seurat_cluster_var,
                title = "CAF markers by Seurat cluster")

DotPlot(seurat_data, features = present_features(seurat_data, caf_markers),
        group.by = "Treatment_Day") &
  theme_classic() & coord_flip() & RotatedAxis() & xlab("") & ylab("")

```

## Lymphoid expression

Lymphoid marker expression by Seurat cluster and by treatment × day.

```{r lymphoid-expression, warning=F, message=F, fig.width = 8, fig.height = max(4, 0.3 * length(present_features(seurat_data, lymph_markers)))}

cluster_dotplot(seurat_data, lymph_markers, group_var = seurat_cluster_var,
                title = "Lymphoid markers by Seurat cluster")

VlnPlot(seurat_data, features = present_features(seurat_data, c("Tigit", "Cd8a", "Cd3e", "Cd4")),
        split.by = "Treatment", group.by = "Day", pt.size = 0, ncol = 1,
        cols = treatment_cols) &
  theme_classic()

```

## Interferon signaling in the final clusters

Mouse IFN gene programs

```{r ifn-genesets, warning=F, message=F}

ifn_groups <- list(
  "Ligands"              = c("Ifnb1", grep("^Ifna[0-9]+$", rownames(seurat_data), value = TRUE),
                             "Ifng", "Ifnl2", "Ifnl3"),
  "Receptors / JAK-STAT" = c("Ifnar1", "Ifnar2", "Ifngr1", "Ifngr2",
                             "Jak1", "Jak2", "Tyk2", "Stat1", "Stat2", "Irf9"),
  "Type I ISGs"          = c("Isg15", "Ifit1", "Ifit2", "Ifit3", "Ifit3b", "Rsad2",
                             "Mx1", "Mx2", "Oas1a", "Oasl1", "Oasl2", "Irf7",
                             "Ifi44", "Ifi27l2a", "Rtp4", "Xaf1", "Bst2", "Zbp1",
                             "Ifih1", "Rigi", "Ddx58", "Isg20"),
  "IFN-g response"       = c("Irf1", "Cxcl9", "Cxcl10", "Gbp2", "Gbp5", "Gbp7",
                             "Igtp", "Iigp1", "Irgm1", "Tgtp1", "Ciita", "Cd274", "Ido1"),
  "Antigen presentation" = c("B2m", "Tap1", "Psmb8", "Psmb9", "H2-K1", "H2-D1",
                             "H2-Ab1", "H2-Aa", "Cd74"),
  "Negative regulators"  = c("Socs1", "Socs3", "Usp18"))

ifn_groups_present <- lapply(ifn_groups, function(g) present_features(seurat_data, g))
ifn_groups_present <- ifn_groups_present[lengths(ifn_groups_present) > 0]
n_spp1_axis <- length(unlist(ifn_groups_present))
n_final_clusters <- length(levels(seurat_data@meta.data[[seurat_cluster_var]]))

# Module scores (require >= 3 panel genes per program)
# Score on the full-resolution assay (all spots), not the sketch assay
score_assay <- "Spatial"   # change if your full assay has another name, e.g. "Spatial.008um"
Assays(seurat_data)
Layers(seurat_data[[score_assay]])

# AddModuleScore needs a single, normalized "data" layer
if (length(Layers(seurat_data[[score_assay]], search = "data")) > 1 ||
    length(Layers(seurat_data[[score_assay]], search = "counts")) > 1) {
  seurat_data[[score_assay]] <- JoinLayers(seurat_data[[score_assay]])
}
if (!"data" %in% Layers(seurat_data[[score_assay]])) {
  seurat_data <- NormalizeData(seurat_data, assay = score_assay, verbose = FALSE)
}

# Keep only program genes that are in this assay and detected in at least one spot
expr_mat <- LayerData(seurat_data, assay = score_assay, layer = "data")
detected <- rownames(expr_mat)[Matrix::rowSums(expr_mat > 0) > 0]

score_sets <- list(IFN_typeI = ifn_groups_present[["Type I ISGs"]],
                   IFN_gamma = ifn_groups_present[["IFN-g response"]],
                   IFN_MHC   = ifn_groups_present[["Antigen presentation"]])
score_sets <- lapply(score_sets, function(g) intersect(g, detected))
score_sets <- score_sets[lengths(score_sets) >= 3]
lengths(score_sets)

for (nm in names(score_sets)) {
  seurat_data <- AddModuleScore(seurat_data, features = list(score_sets[[nm]]),
                                name = nm, assay = score_assay,
                                nbin = 24, ctrl = 100, seed = 1234)
  seurat_data@meta.data[[nm]] <- seurat_data@meta.data[[paste0(nm, "1")]]
  seurat_data@meta.data[[paste0(nm, "1")]] <- NULL
}
ifn_score_cols <- names(score_sets)
rm(expr_mat)

lengths(score_sets)

```

IFN program genes by cluster

```{r ifn-dot-cluster, warning=F, message=F, fig.width = max(8, 0.25 * n_spp1_axis + 3), fig.height = max(4, 0.25 * n_final_clusters + 2)}

ifn_dot <- DotPlot(seurat_data, features = ifn_groups_present,
                   group.by = seurat_cluster_var, cols = c("gray89", "tomato3")) +
  theme_classic() + RotatedAxis() + xlab("") + ylab("") +
  theme(strip.text.x = element_text(angle = 90, hjust = 0, size = 8),
        strip.background = element_blank())
ifn_dot

pdf(file.path(figdir, "VisiumHD_spp1_axis_by_cluster.pdf"),
    width = max(8, 0.25 * n_spp1_axis + 3), height = max(4, 0.25 * n_final_clusters + 2))
ifn_dot
dev.off()

```

IFN module scores by cluster

```{r ifn-score-cluster, warning=F, message=F, fig.width = max(8, 0.35 * n_final_clusters), fig.height = 3 * length(ifn_score_cols)}

VlnPlot(seurat_data, features = ifn_score_cols, group.by = seurat_cluster_var,
        cols = seurat_cluster_col, pt.size = 0, ncol = 1) &
  theme_classic() & RotatedAxis() & NoLegend() & xlab("")

q95 <- sapply(ifn_score_cols, function(s) quantile(seurat_data@meta.data[[s]], 0.95, na.rm = TRUE))

ifn_cluster_summary <- seurat_data@meta.data %>%
  group_by(cluster = .data[[seurat_cluster_var]]) %>%
  dplyr::summarise(n = dplyr::n(),
                   across(all_of(ifn_score_cols), ~ mean(.x, na.rm = TRUE), .names = "{.col}_mean"),
                   across(all_of(ifn_score_cols), ~ mean(.x > q95[cur_column()], na.rm = TRUE),
                          .names = "{.col}_pct_high"),
                   .groups = "drop")

ifn_cluster_summary

```

```{r ifn-score-heatmap, warning=F, message=F, fig.width = max(8, 0.35 * n_final_clusters), fig.height = 3}

ifn_heat_df <- ifn_cluster_summary %>%
  dplyr::select(cluster, ends_with("_mean")) %>%
  pivot_longer(-cluster, names_to = "program", values_to = "score") %>%
  mutate(program = sub("_mean$", "", program)) %>%
  group_by(program) %>%
  mutate(z = as.numeric(scale(score))) %>%
  ungroup()

ifn_heat <- ggplot(ifn_heat_df, aes(x = cluster, y = program, fill = z)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "tomato3",
                       name = "Mean score\n(z across clusters)") +
  theme_classic() + RotatedAxis() + xlab("") + ylab("")

ifn_heat


```

IFN module scores on the tissue

```{r ifn-score-sp, warning=F, message=F, fig.width = 20, fig.height = 8 * length(ifn_score_cols)}

FeaturePlot_scCustom(seurat_object = seurat_data, features = ifn_score_cols,
                     reduction = "sp", num_columns = 1, na_cutoff = NA,
                     raster = TRUE, order = TRUE) &
  coord_fixed() & NoAxes()

```

```{r ifn-score-treatment, warning=F, message=F, fig.width = 16, fig.height = max(6, 2.2 * ceiling(n_final_clusters / 6))}

ifn_sample_df <- seurat_data@meta.data %>%
  filter(as.character(.data[[seurat_cluster_var]]) != "NA") %>%
  group_by(cluster = .data[[seurat_cluster_var]], Sample, Treatment, Day, Treatment_Day) %>%
  dplyr::summarise(n = dplyr::n(),
                   across(all_of(ifn_score_cols), ~ mean(.x, na.rm = TRUE)),
                   .groups = "drop") %>%
  filter(n >= 20)

for (s in ifn_score_cols) {
  p <- ggplot(ifn_sample_df, aes(x = Treatment_Day, y = .data[[s]], fill = Treatment)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.15, size = 0.9) +
    scale_fill_manual(values = treatment_cols) +
    facet_wrap(~ cluster, scales = "free_y", ncol = 6) +
    labs(x = NULL, y = paste0(s, " (per-sample mean)"), title = s) +
    theme_bw(base_size = 10) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p)
  ggsave(file.path(figdir, paste0("VisiumHD_", s, "_by_cluster_treatment.pdf")), p,
         width = 16, height = max(6, 2.2 * ceiling(n_final_clusters / 6)))
}

```

Treatment effect on the IFN scores

```{r ifn-score-stats, warning=F, message=F}

ifn_kw <- ifn_sample_df %>%
  pivot_longer(all_of(ifn_score_cols), names_to = "program", values_to = "score") %>%
  group_by(cluster, program) %>%
  filter(dplyr::n_distinct(Treatment) > 1) %>%
  dplyr::summarise(n_samples = dplyr::n(),
                   p = kruskal.test(score ~ factor(Treatment))$p.value,
                   .groups = "drop") %>%
  group_by(program) %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  ungroup() %>%
  arrange(p_adj)

ifn_kw

#saveRDS(seurat_data, "/scratch/hnatri/VisiumHD_clusters.rds")

```

## Custom annotations

Cluster → cell type labels from the "Visium HD annotations" tab

```{r custom-annot-import, warning=F, message=F}

seurat_data <- readRDS("/scratch/hnatri/VisiumHD_clusters.rds")

gs4_deauth()  # sheet is shared by link; use gs4_auth() instead if it becomes restricted
annot_url <- "https://docs.google.com/spreadsheets/d/1ApwXjEVtpPB87al6q3ab8TKvZYJTh3iNH1cuO-A_OoU/edit?usp=sharing"

custom_annot <- read_sheet(annot_url, sheet = "Visium HD annotations")

seurat_data$celltype <- plyr::mapvalues(seurat_data$new_cluster,
                                  from = custom_annot$cluster,
                                  to = custom_annot$celltype)

celltype_levels <- sort(unique(seurat_data$celltype))
seurat_data$celltype <- factor(seurat_data$celltype, levels = celltype_levels)

celltype_col <- colorRampPalette(brewer.pal(12, "Paired"))(length(celltype_levels))
names(celltype_col) <- celltype_levels

```

Cell types on the tissue

```{r custom-annot-sp, warning=F, message=F, fig.width = 20, fig.height = 8}

p_ct_sp <- DimPlot(seurat_data, reduction = "sp", group.by = "celltype",
                   cols = celltype_col, raster = TRUE) +
  theme_classic() + coord_fixed() + NoAxes()

p_ct_sp

#pdf(file.path(figdir, "VisiumHD_dimplot_celltype.pdf"), width = 20, height = 8)
#p_ct_sp
#dev.off()

```

Cell type composition by treatment × day (pooled spots)

```{r custom-annot-bar, warning=F, message=F, fig.width = 8, fig.height = 5}

create_barplot(seurat_data,
               group_var = "Treatment_Day",
               plot_var = "celltype",
               plot_levels = celltype_levels,
               group_levels = levels(seurat_data$Treatment_Day),
               plot_colors = celltype_col,
               var_names = c("Frequency (%)", ""),
               legend_title = "Cell type")

```

Per-sample proportions. The sample (not the spot) is the unit of replication.
Unannotated spots are excluded from the denominator.

```{r custom-annot-props, warning=F, message=F, fig.width = 14, fig.height = 9}

# Per-sample cell type counts and proportions. Unannotated spots are left out
# of the denominator. Cell types a sample doesn't have are filled in as 0, so
# the GLM and plots include zero counts.
ct_counts <- seurat_data@meta.data %>%
  dplyr::filter(!is.na(celltype), as.character(celltype) != "Unannotated") %>%
  dplyr::mutate(celltype = droplevels(celltype)) %>%
  dplyr::count(Sample, Treatment, Day, Treatment_Day, celltype, name = "n") %>%
  tidyr::complete(tidyr::nesting(Sample, Treatment, Day, Treatment_Day), celltype,
                  fill = list(n = 0)) %>%
  dplyr::group_by(Sample) %>%
  dplyr::mutate(total = sum(n),
                prop  = n / total) %>%
  dplyr::ungroup()

p_ct_prop <- ggplot(ct_counts, aes(x = Treatment_Day, y = prop, color = Treatment)) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5, color = "gray30", linewidth = 0.3) +
  geom_point(size = 1.8, position = position_jitter(width = 0.1, height = 0, seed = 1)) +
  scale_color_manual(values = treatment_cols) +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ celltype, scales = "free_y") +
  labs(x = NULL, y = "Proportion of spots (per sample)") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_ct_prop

#ggsave(file.path(figdir, "VisiumHD_celltype_props_by_group.pdf"), p_ct_prop, width = 14, height = 9)

```

### Day 16: cell type proportion differences between treatments

Per cell type, a quasibinomial GLM on per-sample counts
(cbind(n, total - n) ~ Treatment, Tumor as reference). Models overdispersion
across samples to avoid treating spots as replicates. Omnibus p-values (any treatment
difference) are BH-adjusted across cell types, and per-contrast p-values across all
contrasts.

```{r custom-annot-d16-stats, warning=F, message=F}

d16 <- ct_counts %>%
  dplyr::filter(Day == "16") %>%
  dplyr::mutate(Treatment = factor(Treatment,
                                   levels = intersect(c("Tumor", "Anti-Spp1", "CAR", "Anti-Spp1+CAR"),
                                                      unique(Treatment))))

d16_n <- d16 %>% dplyr::distinct(Sample, Treatment) %>% dplyr::count(Treatment, name = "n_samples")
d16_n

single_groups <- as.character(d16_n$Treatment[d16_n$n_samples < 2])
ref_group <- levels(d16$Treatment)[1]

fit_celltype <- function(df) {
  if (sum(df$n) == 0 || dplyr::n_distinct(df$Treatment) < 2) return(tibble())
  fit  <- glm(cbind(n, total - n) ~ Treatment, family = quasibinomial, data = df)
  if (fit$df.residual < 1) return(tibble())   # no replicated group -> no inference
  null <- glm(cbind(n, total - n) ~ 1, family = quasibinomial, data = df)
  omnibus_p <- anova(null, fit, test = "F")$`Pr(>F)`[2]

  co <- summary(fit)$coefficients
  co <- co[grepl("^Treatment", rownames(co)), , drop = FALSE]
  grp <- sub("^Treatment", "", rownames(co))

  # Groups with zero counts of this cell type give unstable (separated) estimates
  zero_groups <- df %>% dplyr::group_by(Treatment) %>%
    dplyr::summarise(z = sum(n) == 0, .groups = "drop") %>%
    dplyr::filter(z) %>% dplyr::pull(Treatment) %>% as.character()

  tibble(contrast  = paste0(grp, " vs ", ref_group),
         group     = grp,
         log_OR    = co[, "Estimate"],
         OR        = exp(co[, "Estimate"]),
         se        = co[, "Std. Error"],
         p         = co[, "Pr(>|t|)"],
         omnibus_p = omnibus_p,
         resid_df  = fit$df.residual,
         dispersion = summary(fit)$dispersion,
         single_sample_group = grp %in% single_groups | ref_group %in% single_groups,
         zero_count_group    = grp %in% zero_groups | ref_group %in% zero_groups)
}

d16_stats <- d16 %>%
  dplyr::group_by(celltype) %>%
  dplyr::group_modify(~ fit_celltype(.x)) %>%
  dplyr::ungroup()

d16_omnibus <- d16_stats %>%
  dplyr::distinct(celltype, omnibus_p, resid_df, dispersion) %>%
  dplyr::mutate(omnibus_p_adj = p.adjust(omnibus_p, method = "BH")) %>%
  dplyr::arrange(omnibus_p_adj)

d16_stats <- d16_stats %>%
  dplyr::mutate(p_adj = p.adjust(p, method = "BH")) %>%
  dplyr::left_join(d16_omnibus %>% dplyr::select(celltype, omnibus_p_adj), by = "celltype") %>%
  dplyr::arrange(omnibus_p_adj, p_adj)

d16_omnibus

d16_stats

```

Speckle/propeller

```{r custom-annot-d16-propeller, warning=F, message=F}

md16 <- seurat_data@meta.data %>%
  dplyr::filter(as.character(Day) == "16", celltype != "Unannotated")
d16_propeller <- speckle::propeller(clusters = droplevels(md16$celltype),
                                    sample   = as.character(md16$Sample),
                                    group    = as.character(md16$Treatment),
                                    transform = "logit")
d16_propeller

```

Day 16 per-sample proportions with omnibus FDR

```{r custom-annot-d16-plot, warning=F, message=F, fig.width = 12, fig.height = 9}

facet_labs <- d16_omnibus %>%
  dplyr::mutate(lab = paste0(celltype, "\nFDR = ", signif(omnibus_p_adj, 2))) %>%
  { setNames(.$lab, as.character(.$celltype)) }

p_d16 <- ggplot(d16, aes(x = Treatment, y = prop, color = Treatment)) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5, color = "gray30", linewidth = 0.3) +
  geom_point(size = 2, position = position_jitter(width = 0.1, height = 0, seed = 1)) +
  scale_color_manual(values = treatment_cols) +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ celltype, scales = "free_y",
             labeller = labeller(celltype = function(x) ifelse(x %in% names(facet_labs),
                                                               facet_labs[x], x))) +
  labs(x = NULL, y = "Proportion of spots (per sample)",
       title = "Day 16 cell type proportions",
       subtitle = if (length(single_groups))
         paste0("Single-sample group(s): ", paste(single_groups, collapse = ", ")) else NULL) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
p_d16

#ggsave(file.path(figdir, "VisiumHD_celltype_props_D16.pdf"), p_d16, width = 12, height = 9)

```
